Contracts for unions and intersections are hard for Nickel

### 2021-05-06 paper.tex
Move to the ACMart 10pt format
Use latest acmart
Some comments
+typos

\documentclass{article}
\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{hyperref}

@@

  \usepackage{xargs}
  \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
  % ^^ Need for pgfsyspdfmark apparently?
  \ifx\noeditingmarks\undefined
      % \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
      \newcommand{\Red}[1]{{\color{red}{#1}}}
      \newcommand{\newaudit}[1]{{\color{blue}{#1}}}
      \newcommand{\note}[1]{{\color{blue}{\begin{itemize} \item {#1} \end{itemize}}}}
      \newenvironment{alt}{\color{red}}{}

@@

\usepackage{biblatex}
\addbibresource{nickel.bib}

@@

Build systems, package managers, operating systems, cloud infrastructures, and
web services are examples of modern complex software systems that require an
extensive and non-trivial configuration in order to make them adapted to each
different use-case. To manage the growing complexity that is then offloaded to
configurations, the declarative approach has become more and more popular,
illustrated for example by the infrastructure-as-code paradigm. This leads
configuration to play an important role in critical aspects of software
engineering, including security, availability, and maintainability.

However, static text-based configuration alone is falling short of expressivity,
and is seldom sufficient. Data cannot be transformed, combined nor shared,
resulting at best in boilerplate and duplication of information, or at worst, in
data being invalid or inconsistent. Data validation is not supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
done. Correctly configuring a modern system is hard and failures may have
substantial negative consequences.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts. In "first-class contracts", we include the
ability to write user-defined contracts. Because built-in types are too
simplistic for the purpose of data validation, the ability of writing a small
validation function for, say, an URL, and having it immediately available as a
contract that can be composed with other type constructors, is in our
opinion one of the main appeal of Nickel contracts system. In this hybrid
system, each type constructor of the static type system - arrows, records,
foralls - must have a contract counterpart that checks at run-time that a value
is indeed a member of the corresponding type. Contracts for various extensions
of the simply typed $\lambda$-calculus have been investigated \cite{BlameForAll,
    KeilThiemannUnionIntersection, RootCauseOfBlame, DependentContracts,
GradualTypingClasses}, with solutions of varying complexity.  Amongst all the
common extensions to the polymorphic lambda calculus, one of the most useful and
widespread turns out to be one of the most challenging: \emph{union types}.

@@

In statically typed languages, and more specifically in functional programming
languages, unions are usually implemented as algebraic data types [references],
also called \emph{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

@@

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors and destructors
with well-defined semantics. In contrast, in a gradually typed language like
Nickel, the question of the representation of tagged unions as untyped values
arises.

@@

Nickel does not feature dependent types, which would open a whole new level of
complexity. Alas, choosing the non dependent representation as a default is not
fully satisfying.  First, it does not map precisely to a JSON value: in
consequence, a lot of illegal combinations are structurally valid as untyped
JSON values. But this is to be expected of any representation. More importantly,
this does not necessarily match with the configuration schemas one can find in
the wild, that favor simple, untagged union.

In dynamically typed languages, it is customary to directly store different kind
of data in one variable without further ceremony. The runtime system needs to
attach type tags to values anyway, which can usually be observed using functions
like JavaScript's \verb+typeof+. Using so-called \emph{untagged unions} has
already been an idiom in dynamically typed language for a long time
[ref/examples], and being able to type and handle these unions is an important
aspect of the migration to a gradual type system. Indeed, a lot of gradually
typed languages either existed originally as dynamically typed language or aim
at backward compatibility with a dynamically typed base language. This is why
they tend to favor supporting untagged union types, the typed counterpart of
this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

@@

\begin{enumerate}
    \item Most of the encountered issues already appear with bare unions.
    \item Intersection being dual to unions, they are showing up in some way
        when unions are put in negative positions. For example, a contract for
        $(A \cup B) \to C$ in a system without intersections must be equivalent
        to the contract $(A \to C) \cap (B \to C)$
\end{enumerate}

@@

Among those contracts, they provide \texttt{and/c} and \texttt{or/c} contracts.
These contracts are lightweight versions of unions and intersections, they simply
check that every contract holds, in the case of and/c, and that at least one contract
holds, in the case of or/c.

So, for instance, the usual usecase of intersection to encode overloading
does not work on Racket. Take a look at the next example, the function
\texttt{overload} shoudl accept a call like \texttt{(overload 3)}, but it fails
since it 3 is not a valid \texttt{string?}.

@@

Even worst, the or/c contract doesn't allow to have multiple higher order
contracts that could (potentially) apply to a function, these have to be
differentiated by first order markers (for instance, the number of arguments).

@@

\printbibliography

@@

>>>>

\documentclass[sigplan,10pt]{acmart}
%\usepackage{amssymb}
\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{hyperref}

@@

  \usepackage{xargs}
  \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
  % ^^ Need for pgfsyspdfmark apparently?
  \ifx\noeditingmarks\undefined
      \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
      \newcommand{\Red}[1]{{\color{red}{#1}}}
      \newcommand{\newaudit}[1]{{\color{blue}{#1}}}
      \newcommand{\note}[1]{{\color{blue}{\begin{itemize} \item {#1} \end{itemize}}}}
      \newenvironment{alt}{\color{red}}{}

@@

\bibliographystyle{ACM-Reference-Format}

@@

Build systems, package managers, operating systems, cloud infrastructures, and
web services\unsure{Also continuous integration} are examples of modern complex software systems that require an
extensive and non-trivial configuration in order to make them adapted to each
different use-case. To manage the growing complexity that is then offloaded to
configurations, the declarative approach has become more and more popular,
illustrated for example by the infrastructure-as-code paradigm. This leads
configuration to play an important role in critical aspects of software
engineering, including security, availability, and maintainability.

However, static text-based configuration alone is falling short of expressiveness,
and is seldom sufficient. Data cannot be transformed, combined nor shared,
resulting at best in boilerplate and duplication of information, or at worst, in
data being invalid or inconsistent. Data validation is not supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
done. Correctly configuring a modern system is hard and failures may have
substantial negative consequences.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Felleisen introduce a principled approach to run-time assertion checking that
nicely supports higher-order functions and introduces the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \&
  Felleisen}.\unsure{The historical bit should probably be moved to
  related works eventually} Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts. In "first-class contracts", we include the
ability to write user-defined contracts. Because built-in types are too
simplistic for the purpose of data validation, the ability of writing a small
validation function for, say, a URL, and having it immediately available as a
contract that can be composed with other type
constructors\unsure{Needs code to illustrate.}, is in our
opinion one of the main appeal of Nickel's contracts system. In this hybrid
system, each type constructor of the static type system - arrows, records,
foralls - must have a contract counterpart that checks at run-time that a value
is indeed a member of the corresponding type. Contracts for various extensions
of the simply typed $\lambda$-calculus have been investigated \cite{BlameForAll,
    KeilThiemannUnionIntersection, RootCauseOfBlame, DependentContracts,
GradualTypingClasses}, with solutions of varying complexity.  Amongst all the
common extensions to the polymorphic lambda calculus, one of the most useful and
widespread turns out to be one of the most challenging: \emph{union types}.

@@

In statically typed languages, and more specifically in functional programming
languages, unions are usually implemented as algebraic data
types\unsure{This is a very misleading statement at best. Sums are
  "tagged union" sure, but that refers to an encoding a sums with
  unions, certainly not an encoding of unions with sums!} [references],
also called \emph{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

@@

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a first-class data type associated with constructors and destructors
with well-defined semantics. In contrast, in a gradually typed language like
Nickel, the question of the representation of tagged unions as untyped values
arises.

@@

Nickel does not feature dependent types, which would open a whole new level of
complexity. Alas, choosing the non dependent representation as a default is not
fully satisfying.  First, it does not map precisely to a JSON value: in
consequence, a lot of illegal combinations are structurally valid as untyped
JSON values. But this is to be expected of any representation. More importantly,
this does not necessarily match with the configuration schemas one can find in
the wild, which favor simple, untagged union.

In dynamically typed languages, it is customary to directly store different kinds
of data in one variable without further ceremony. The runtime system needs to
attach type tags to values anyway, which can usually be observed using functions
like JavaScript's \verb+typeof+. Using so-called \emph{untagged
  unions}\unsure{If you can observe with typeof, then it's
  pretty much a tagged union, though the tag is implicit} has
already been an idiom in dynamically typed language for a long time
[ref/examples], and being able to type and handle these unions is an important
aspect of the migration to a gradual type system. Indeed, a lot of gradually
typed languages either existed originally as dynamically typed language or aim
at backward compatibility with a dynamically typed base language. This is why
they tend to favor supporting untagged union types, the typed counterpart of
this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

@@

\begin{enumerate}
    \item Most of the encountered issues already appear with bare unions.
    \item Intersection being dual to unions, they are showing up in some way
        when unions are put in negative positions. For example, a contract for
        $(A \cup B) \to C$ in a system without intersections must be
        equivalent\unsure{``Must be'' is a rather strong statement, we
        will find a way to tone it down}
        to the contract $(A \to C) \cap (B \to C)$
\end{enumerate}

\unsure{Some bits missing so far: we are interested in higher-order
  contracts, Nickel is a higher-order \& pure language, contracts as
  functions. We've said that user-defined contracts are a thing, it
  probably matters so we may have to hammer it down.}

\unsure{Something that we must mention somewhere: if we want to check
  that something is a list of As lazily (to avoid making functions
  like hd be O(n), for instance), then we have a lot of the problems of
  higher-order contracts anyway.}

@@

Among those contracts, they provide \texttt{and/c} and \texttt{or/c} contracts.
These contracts are lightweight versions of unions and intersections, they simply
check that every contract holds, in the case of and/c, and that at least one contract
holds, in the case of or/c.\unsure{It doesn't really make sense for a
  contract to “hold”, in general}

So, for instance, the usual usecase of intersection to encode overloading
does not work on Racket. Take a look at the next example, the function
\texttt{overload} should accept a call like \texttt{(overload 3)}, but it fails
since it 3 is not a valid \texttt{string?}.

@@

Even worst, the or/c contract doesn't allow to have multiple higher order
contracts that could (potentially) apply to a function, these have to be
differentiated by first order markers (for instance, the number of
arguments).\unsure{This section is overall a bit hard to follow. I
  think I've figured out what it means, but it needs more love. Also
  give an example which doesn't work and one that does for or/c.}

@@

\unsure{We ought to group all the static typing in one subsection}

\bibliography{nickel}

@@

% LocalWords:  Felleisen
